[Re] Chaos in a long-term experiment with a plankton community

### 2016-02-17 article/article.tex
small edits

    \usepackage{authblk}
                        \author[1]{Owen Petchey}
                    \author[1]{Marco Plebani}
                    \author[1]{Frank Pennekamp}
                            \affil[1]{Institute of Evolutionary Biology and Environmental Studies, University
of Zurich, Zurich, Switzerland}

@@

    \vspace{-1mm}
  \begin{repobox}
    \bfseries\normalsize
      \href{http://github.com/rescience/rescience-submission/article}{\faGithubAlt~Article repository}
  \end{repobox}
      \vspace{-1mm}
  \begin{repobox}
    \bfseries\normalsize
      \href{http://github.com/rescience/rescience-submission/code}{\faGithubAlt~Code repository}
  \end{repobox}
        \hypersetup{urlcolor=blue}
}

@@

The original paper describes analyses of fluctuations in the abundance
of organisms in a plankton community derived from the Baltic Sea, housed
in a laboratory environment. The length of the time series (samples
every few days for 2,300 days) allowed for analyses revealing that the
observed dynamics exhibited characteristics consistent with chaos
produced by non-linear species interactions. The article concludes that
stability is not required for persistence of complex food webs, and that
long-term prediction of abundances may be fundamentally impossible. The
demonstration of chaotic dynamics and limited forecast horizons (sensu
\textcite{Petchey2015}) is important in the field of ecology, since the
ability to predict dynamics is an open question with considerable
applied importance \textcite{Petchey2015} \textcite{Mouquet2015}.

@@

This reproduction started with the raw data (source given below) and
used information from the original paper, the supplementary information
\href{http://www.nature.com/nature/journal/v451/n7180/extref/nature06512-s1.pdf}{the
Supplement to the Nature paper}, and communications with Elisa Benincà,
who provided for comparison digitised data from the original article,
and Stephen Ellner, who provided code and data used to produce results
in the original paper. Use of this data and code in this reproduction is
indicated below.

@@

The data are available as an Excel file supplement to
\href{http://onlinelibrary.wiley.com/doi/10.1111/j.1461-0248.2009.01391.x/abstract}{an
Ecology Letters publication} \textcite{Beninca2009}. The Excel file
contains several datasheets. Two were particularly important, as they
were the source of the raw data (one contains original species
abundances, the other the nutrient concentrations). We saved these two
datasheets, without any alteration, as comma separated value (csv) text
files. In the code associated with this reproduction, these data files
are read from the associated github repository. Another datasheet
contained transformed data (we also saved this as csv file, in order to
use it in this reproduction). We also received a dataset direct from
Steve Ellner, see below for details.

The original abundance dataset contains a duplicated date: 28/10/1996
(row 709 and 710 in excel sheet). We changed this (in R) to 26/10/1996,
making the observation occur half way between the two surrounding dates.
The original Protozoa variable in the original dataset contains some
numbers with a comma (instead of period) as the decimal separator; these
were changed to periods.

@@

The R language and environment for statistical computing and graphics
was used to make the reproduction. Additional R packages required are
specified in the code associated with this reproduction. The code for
making the analyses and figures presented in this reproduction resides
in an R markdown document, as well as a source file containing some
required functions. Some of the code takes several minutes to run, so an
intermediate data file is provided with results from this code.

@@

The reproduced populations dynamics were very similar to those in figure
1b-g of the original publication (figure \ref{fig:dynamics}). As most
analyses were performed on fourth root transformed values, we plot
these, rather than raw abundances with a y-axis break as in the original
article.

@@

The original publication states that long sequences of zeros were
removed from time series prior to transformation and further analysis.
We removed zeros by matching the removals to those in the transformed
version of the data in the published Excel file mentioned above.
Transformation steps were:

@@

Correlations among species abundances presented in Table 1 of the
original article closely matched our reproduced correlations, calculated
from the transformed data with zeros removed (figure
\ref{fig:corr_comp}). Deviations between the original and reproduced
correlations are relatively small (less than 0.072 units) and
infrequent.

@@

Reproduced divergence rates (figure \ref{fig:divergence}) and therefore
Lyapunov exponents (figure \ref{fig:LE_comparison}) were somewhat
different from those in the original article. The original article
states: ``the distance between initially nearby trajectories increased
over time, and reached a plateau after about 20--30 days''. The
reproduced results are consistent with this statement. The original
article also stated that the analyses ``yielded significantly positive
Lyapunov exponents of strikingly similar value for all species (Fig. 3;
mean exponent = 0.057 per day, s.d. = 0.005 per day, n = 9)''.
Reproduced exponents had very similar mean value, but had larger
standard deviation (mean = 0.060 and s.d. = 0.014).

\begin{figure}[htbp]
\centering
\includegraphics{figures/div_rate.pdf}
\caption{\label{fig:divergence}Reproduced divergence rates and Lyapunov
exponents (figure 3 in the original article).}
\end{figure}

@@

The original paper reported a global Lyapunov exponent calculated via
two modelling approaches (neural network and generalised additive models
{[}GAMs{]}). Only the GAM approach was reproduced, with the assistance
of code donated by Stephen Ellner. The original article obtained a
global Lyapunov exponent of λ=0.08 day-1. The reproduced value was 0.04.
We did not reproduce the bootstrapping used to give confidence intervals
around this estimate.

@@

The reproduced results were qualitatively identical to those in the
original article, and therefore support the conclusion of chaotic
dynamics produced by non-linear species interactions. For example, all
Lyapunov exponents estimated by direct method were positive, as in the
original article, consistent with chaotic dynamics. Quantitative
differences may have resulted from difference in algorithms used. For
example,the original used the
\href{http://www.mpipks-dresden.mpg.de/~tisean/}{Tisean software} to
calculate Lyapunov exponents. As this was available from CRAN
\href{http://cran.r-project.org/web/packages/RTisean/index.html}{until
mid 2014} and since it is a bit less well integrated with R, we instead
use the tseriesChaos package \textcite{tseriesChaos}, which in any case
was largely inspired by the TISEAN project. In addition, there may have
been some difference in algorithm parameters, as not all parameters
required by the functions we used were reported in the original ms.

Considerable quantitative difference in predictability decay
(\ref{fig:prediction_distance}) remain unexplained, though could result
from original analyses using neural networks, and the reproduction using
GAMs.

@@

This reproduction was made as part of the Reproducible Research in
Ecology, Evolution, Behaviour, and Environmental Studies (RREEBES)
Course, lead by Owen Petchey at the University of Zurich. More
information about the course
\href{https://github.com/opetchey/RREEBES/blob/master/README.md}{here}
on github. Many thanks to Elisa Benincà and Steve Ellner for assistance,
and to Steve for providing unpublished data and code.

>>>>

    \usepackage{authblk}
                        \author[1]{Owen Petchey}
                    \author[1]{Marco Plebani}
                    \author[1]{Frank Pennekamp}
                            \affil[1]{Department of Evolutionary Biology and Environmental Studies, University
of Zurich, Zurich, Switzerland}

@@

    \vspace{-1mm}
  \begin{repobox}
    \bfseries\normalsize
      \href{https://github.com/opetchey/ReScience-submission/tree/petchey-plebani-pennekamp-2016/article}{\faGithubAlt~Article repository}
  \end{repobox}
      \vspace{-1mm}
  \begin{repobox}
    \bfseries\normalsize
      \href{https://github.com/opetchey/ReScience-submission/tree/petchey-plebani-pennekamp-2016/code}{\faGithubAlt~Code repository}
  \end{repobox}
      \vspace{-1mm}
  \begin{repobox}
    \bfseries\normalsize
      \href{https://github.com/opetchey/ReScience-submission/tree/petchey-plebani-pennekamp-2016/data}{\faGithubAlt~Data repository}
  \end{repobox}
      \hypersetup{urlcolor=blue}
}

@@

The original paper describes analyses of fluctuations in the abundance
of organisms in a plankton community derived from the Baltic Sea, housed
in a laboratory environment. The length of the time series (samples
every few days for 2,300 days) allowed for analyses revealing that the
observed dynamics exhibited characteristics consistent with chaos
produced by non-linear species interactions. The article concludes that
stability is not required for persistence of complex food webs, and that
long-term prediction of abundances may be fundamentally impossible. The
demonstration of chaotic dynamics and limited forecast horizons (sensu
\textcite{Petchey2015}) is important in the field of ecology, since the
ability to predict dynamics is an open question with considerable
applied importance (\textcite{Petchey2015} \textcite{Mouquet2015}.)

@@

This reproduction started with the raw data (source given below) and
used information from the original paper,
\href{http://www.nature.com/nature/journal/v451/n7180/extref/nature06512-s1.pdf}{the
supplement to the original paper}, and communications with Elisa
Benincà, who provided for comparison digitised data from the original
article, and Stephen Ellner, who provided code and data used to produce
results in the original paper. Use of this data and code in this
reproduction is indicated below.

@@

The data are available as an Excel file supplement to
\href{http://onlinelibrary.wiley.com/doi/10.1111/j.1461-0248.2009.01391.x/abstract}{an
Ecology Letters publication} (\textcite{Beninca2009}). The Excel file
contains several datasheets. Two were particularly important, as they
were the source of the raw data (one contains original species
abundances, the other the nutrient concentrations). We saved these two
datasheets, without any alteration, as comma separated value (csv) text
files. In the code associated with this reproduction, these data files
are read from the associated github repository. Another datasheet
contained transformed data (we also saved this as csv file, in order to
use it in this reproduction). We also received a dataset direct from
Steve Ellner, see below for details.

The original abundance dataset contains a duplicated date: 28/10/1996
(row 709 and 710 in excel sheet). We changed this (in R) to 26/10/1996,
making the observation occur half way between the two surrounding dates.
The Protozoa variable in the original dataset contains some numbers with
a comma (instead of period) as the decimal separator; these were changed
to periods.

@@

The R language and environment for statistical computing and graphics
(\textcite{R}) was used to make the reproduction. Additional R packages
required are specified in the code associated with this reproduction.
The code for making the analyses and figures presented in this
reproduction resides in an R markdown document, as well as a source file
containing some required functions. Some of the code takes several
minutes to run, so an intermediate data file is provided with results
from this code.

@@

The reproduced populations dynamics (figure \ref{fig:dynamics}) were
very similar to those in figure 1b-g of the original publication. As
most analyses were performed on fourth root transformed values, we plot
these, rather than raw abundances with a y-axis break as in the original
article.

@@

The original publication states that long sequences of zeros were
removed from time series prior to transformation and further analysis.
We removed zeros by matching the removals to those in the transformed
version of the data in the published Excel file mentioned above.
Subsequent transformation steps were:

@@

Correlations among species abundances presented in Table 1 of the
original article closely matched our reproduced correlations, calculated
from the transformed data with zeros removed (figure
\ref{fig:corr_comp}). Deviations between the original and reproduced
correlations are relatively small and infrequent.

@@

The original article states: ``the distance between initially nearby
trajectories increased over time, and reached a plateau after about
20--30 days''. The reproduced results (figure \ref{fig:divergence}) are
consistent with this statement. The original article also stated that
the analyses ``yielded significantly positive Lyapunov exponents of
strikingly similar value for all species (Fig. 3; mean exponent = 0.057
per day, s.d. = 0.005 per day, n = 9)''. Reproduced exponents had very
similar mean value, but had larger standard deviation (mean = 0.060 and
s.d. = 0.014), resulting from quantitative differences in reproduced
divergence rates (figure \ref{fig:divergence}) and therefore Lyapunov
exponents (figure \ref{fig:LE_comparison}).

\begin{figure}[htbp]
\centering
\includegraphics{figures/div_rate.pdf}
\caption{\label{fig:divergence}Reproduced divergence rates and Lyapunov
exponents (equivalent to figure 3 in the original article).}
\end{figure}

@@

The original paper reported a global Lyapunov exponent calculated via
two modelling approaches (neural network and generalised additive models
{[}GAMs{]}). Only the GAM approach was reproduced, with the assistance
of code donated by Stephen Ellner. The original article obtained a
global Lyapunov exponent of 0.08 per day. The reproduced value was 0.04.
We did not reproduce the bootstrapping used to give confidence intervals
around this estimate.

@@

The reproduced results were qualitatively identical to those in the
original article, and therefore support the conclusion of chaotic
dynamics. For example, all Lyapunov exponents estimated by direct method
were positive, as in the original article. Quantitative differences may
have resulted from difference in algorithms used. For example, the
original used the
\href{http://www.mpipks-dresden.mpg.de/~tisean/}{Tisean software} to
calculate Lyapunov exponents. As this was available from CRAN
\href{http://cran.r-project.org/web/packages/RTisean/index.html}{until
mid 2014} and since it is a bit less well integrated with R, we instead
used the tseriesChaos package (\textcite{tseriesChaos}), which in any
case was largely inspired by the TISEAN project. In addition, there may
have been some difference in algorithm parameters, as not all parameters
required by the function we used were reported in the original ms.

Considerable quantitative difference in predictability decay (figure
\ref{fig:prediction_distance}) remain unexplained, though could result
from original analyses using neural networks, and the reproduction using
GAMs.

@@

This reproduction was made as part of the Reproducible Research in
Ecology, Evolution, Behaviour, and Environmental Studies (RREEBES)
Course, lead by Owen Petchey at the University of Zurich. More
information about the course
\href{https://github.com/opetchey/RREEBES/blob/master/README.md}{here}
on github. Many thanks to Elisa Benincà and Stephen Ellner for
assistance, and to Stephen Ellner for providing unpublished data and
code.
